---
title: "Data handling"
output:
  pdf_document: default
  html_notebook: default
---
# Data Science Basics: Reading and writing files

## 1. Reading a table

This is code that was shown to you in the slides.

We explicitely import the readr library, then use its read_csv function, that returns a tibble data frame:

```{r}
library("readr")
df <- readr::read_csv("../data/chains.csv")
head(df)
```

Just typing the name of the data frame will show its content:
```{r}
df
```

## Indexing

We can also access a specific column:

```{r}
df$length
```

There are different ways to access a column:

```{r}
df$`PDB code`
with(df, `PDB code`)
df[["PDB code"]]
```

Compare the next output with the previous example! This time, we will get a dataframe back.
```{r}
df["PDB code"]
```


Accessing a row:
```{r}
df[1,]
```

Or multiple rows:
```{r}
df[1:3,]
```

Or specific rows (note the absence of row #2 in the output):
```{r}
df[c(1,3,4),]
```

Combining row and column indexing:
```{r}
df[1:2,2:3]
```


```{r}
df[1:2,"PDB code"]
```


```{r}
df[1:2,c("PDB code", "chain letter")]
```

