---
title: "ds_17_viz"
output:
  pdf_document: default
  html_document: default
date: "2023-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will read some protein data:
```{r}
library(readr)
df <- readr::read_csv("../Resources_practicals/data/data/proteins.csv")
df$n_residues <- as.numeric(df$n_residues)
df$resolution <- as.numeric(df$resolution)
```


We may now wonder what organisms those proteins come from.
We will use the `count` function from `dplyr` for that:

```{r}
library(dplyr)
source.df <- count(df, source)
source.df
```

There are a great many species, looking through it by hand may not be very practical....
Let's plot our data!

We will use `ggplot` for that. GGplot uses its own syntax
```{r}
library(ggplot2)
ggplot(df, aes(x = source)) +
  geom_bar()
```

This is not very helpful! We would like to only plot the most common species for now.
We will order the data by counts, and keep the first five, then plot as before:
```{r}
library(forcats)
source.df[order(source.df$n, decreasing = TRUE), ] %>%
  filter(!is.na(source)) %>% # data cleaning
  top_n(5) %>%
  ggplot( aes(x=source, y=n)) +
  geom_bar(stat = 'identity')
```
Much better! We can see the most common species easily. One of them is not like the others... Some data cleaning might be in order.
N.B: We could also use the `forcats` library to do that (you can always write custom transforms manually, when needed); you can find a nice overview of forcats methods at this address: https://raw.githubusercontent.com/rstudio/cheatsheets/main/factors.pdf

Another way to group and filter our data, using `dplyr`, and starting from the original dataframe (this time we will keep all species present more than 500 times):
```{r}
df %>%
  group_by(source) %>%
  filter(n() > 500) %>%
  ggplot(aes(x=source)) +
  geom_bar()
```
Note that the code is easier to follow, each step is cleanly separated from the next.
This will make it easier to change it in the future, or to find out a problem along the way!



Let's now see some other statistics on our data, for example the number of residues per proteins, for the top species:
```{r}
df %>%
  group_by(source) %>%
  filter(!is.na(source)) %>% # data cleaning
  filter(n() > 500) %>%
  filter(n_residues < 1000) %>%
  ggplot(aes(x=source, y=n_residues, fill=source)) +
  geom_boxplot()
```

Different plots will show us different views of our data:
```{r}
df %>%
  group_by(source) %>%
  filter(!is.na(source)) %>% # data cleaning
  filter(n() > 500) %>%
  filter(n_residues < 1000) %>%
  ggplot(aes(x=source, y=n_residues, fill=source)) +
  geom_violin()
```

```{r}
df
```

What method has the lowest average resolution? The highest?
Try to fix the plot to figure it out:
```{r}
df %>%
  group_by(exp_method) %>%
  filter(resolution < 100) %>% #something is missing here...
  ggplot(aes(x=exp_method, y=resolution, fill=exp_method)) +
    geom_boxplot()
```


We can observe trends, assuming our data is correctly filtered:
```{r}
df %>%
  filter(n_residues < 1000) %>%
  filter(resolution < 5) %>%
  ggplot(aes(x=n_residues, y=resolution)) +
  geom_point()
```
Bonus question: why are there horizontal lines in this plot? Tip: We all like round numbers...

Fairly complex plots can be created rather quickly, using methods from ggplot, (and here, `ggExtra`)
```{r}
library(ggExtra)
p <- df %>%
  filter(n_residues < 1000) %>%
  filter(resolution < 5) %>%
  ggplot(aes(x=n_residues, y=resolution, size=n_chains)) +
  geom_point()
ggMarginal(p, type="histogram")
```


In turn, this type of plot can be easily packaged together in a Rshiny app,
a local app that provides interactive controls for your plot:
(N.B: example taken from: https://shiny.posit.co/)
```{r}
library(shiny)
library(bslib)
library(dplyr)
library(ggplot2)
library(ggExtra)

penguins_csv <- "https://raw.githubusercontent.com/jcheng5/simplepenguins.R/main/penguins.csv"

df <- readr::read_csv(penguins_csv)
# Find subset of columns that are suitable for scatter plot
df_num <- df |> select(where(is.numeric), -Year)

ui <- page_sidebar(
  theme = bs_theme(bootswatch = "minty"),
  sidebar = sidebar(
    varSelectInput("xvar", "X variable", df_num, selected = "Bill Length (mm)"),
    varSelectInput("yvar", "Y variable", df_num, selected = "Bill Depth (mm)"),
    checkboxGroupInput(
      "species", "Filter by species",
      choices = unique(df$Species), 
      selected = unique(df$Species)
    ),
    hr(), # Add a horizontal rule
    checkboxInput("by_species", "Show species", TRUE),
    checkboxInput("show_margins", "Show marginal plots", TRUE),
    checkboxInput("smooth", "Add smoother"),
  ),
  plotOutput("scatter")
)

server <- function(input, output, session) {
  subsetted <- reactive({
    req(input$species)
    df |> filter(Species %in% input$species)
  })
  
  output$scatter <- renderPlot({
    p <- ggplot(subsetted(), aes(!!input$xvar, !!input$yvar)) + list(
      theme(legend.position = "bottom"),
      if (input$by_species) aes(color = Species),
      geom_point(),
      if (input$smooth) geom_smooth()
    )
    
    if (input$show_margins) {
      margin_type <- if (input$by_species) "density" else "histogram"
      p <- ggExtra::ggMarginal(p, type = margin_type, margins = "both",
                               size = 8, groupColour = input$by_species, groupFill = input$by_species)
    }
    
    p
  }, res = 100)
}

shinyApp(ui, server)
```


Pratical question HLA-A*02 dataset

This is a dataset of **T-cell epitopes** taken from the VDJdb database. You don't need to understand the biology here, although to make minimal sense of this data it is worth knowing that:

- Each row contains a short protein sequence (called an epitope) from a source species and a corresponding sequence (the CDR3) from a T-cell receptor (T cells are one of the key types of cell within our adaptive immune systems)
- Presentation of epitopes occurs on the surface of our regular cells and is carried out by special molecules, some of which are encoded by HLA-A*02 genes. 


**1. Species counts**

Draw a barplot (with an appropriate colour palette) showing species (column `Epitope species`) on the y-axis and the number of rows on the x-axis (this represents the unique number of TCR/epitope combinations). Exclude species that have fewer than 50 rows. 

```{r}
epitope_fname = '../Resources_practicals/data/data/HLA-A2.tsv'

epitope <- read_tsv(epitope_fname)

epitope %>%
  group_by(`Epitope species`) %>%
  summarise(n = n()) %>%
  arrange(n) %>%
  filter(n > 50) %>%
  ggplot(aes(x=n, y=`Epitope species`, fill=n)) +
  scale_color_viridis_b() +
  geom_bar(stat="identity")
```

**2. Plotting epitope length and CDR3 length**

Create a single plot containing two histograms: one for epitope lengths and one for CDR3 lengths. 

- Exclude data to focus in on the area of interest as appropriate
- Add a legend.

```{r}
epitope %>%
  mutate (epitope_length = nchar(Epitope)) %>%
  mutate (CDR3_length = nchar(CDR3)) %>%
  select (epitope_length, CDR3_length) %>%
  pivot_longer (cols = c(epitope_length, CDR3_length)) %>%
  #filter(name %in% "epitope_length") %>%
  ggplot(aes(x=value, fill=name)) +
  geom_histogram(bins = 35, alpha=0.6, position="identity")

```


3** (Optional, hard!) Create a Shiny app

Can you create an interactive view of this dataset?
Start from the shiny code above.

```{r}
library(shiny)
library(bslib)
library(dplyr)
library(ggplot2)
library(ggExtra)

epitope <- read_tsv(epitope_fname)

# modify the dataset to select data that will generate the histograms
epitope %>%
  mutate (epitope_length = nchar(Epitope)) %>%
  mutate (CDR3_length = nchar(CDR3)) %>%
  select (epitope_length, CDR3_length) %>%
  pivot_longer (cols = c(epitope_length, CDR3_length)) -> df

ui <- page_sidebar(
  theme = bs_theme(bootswatch = "minty"),
  sidebar = sidebar(
    checkboxGroupInput(
      "nom", 
      "Filter by protein sequence",
      choices = unique(df$name), 
      selected = unique(df$name)
    ),
    hr(), # Add a horizontal rule
  ),
  plotOutput("histo")
)

server <- function(input, output, session) {
  subsetted <- reactive({
    req(input$nom)
    df |> filter(name %in% input$nom)
  })
  
  output$histo <- renderPlot({
    p <- ggplot(subsetted(), aes(value, fill=name)) + 
         geom_histogram(bins = 35, alpha=0.6, position="identity") +
         if (length(input$nom) > 1) {
           scale_fill_manual(values=c("#69b3a2", "#404080")) 
         } else if (length(input$nom) == 1 & "epitope_length" %in% input$nom) {
           scale_fill_manual(values="#404080") 
         } else if (length(input$nom) == 1 & "CDR3_length" %in% input$nom) {
           scale_fill_manual(values="#69b3a2") 
         }
    p
  }, 
  res = 100)
}

shinyApp(ui, server)

```

